---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

# AgriPatterns_Natura2000

<!-- badges: start -->
<!-- badges: end -->

The goal of AgriPatterns_Natura2000 is to document the analyses made for the manuscript entitled "Ecological principles for designating areas for afforestation that co-benefit biodiversity", Specifically, this repo explains how the proportions of agriculture in Natura 2000 nature protection areas in different countries of the EU.

## Loading necessary package

First we need to load the necessary packages

```{r loadpackages}
library(tidyverse)
library(readr)
library(readxl)
library(forcats)
library(stringr)
library(terra)
library(readxl)
library(RColorBrewer)
```



## Data origin and procedure

Corine level 3 was retrieved from [European Environmental Agency](https://www.eea.europa.eu/en/analysis/maps-and-charts/natura-2000-data-viewer-dashboards#:~:text=The%20interactive%20map%20in%20the%20fifth%20page%20shows%20the%20percentage). 


# loading data and calculating the percentage proportion of each type of agriculture in each country

```{r}
#Loading data
level3 <- read_excel("level3.xlsx")

natura2000_area <- level3 |>  
  mutate(natura2000_area = rowSums(across(where(is.numeric)), na.rm = T)) |>  
  distinct(country, .keep_all = T) |> 
  select(country, natura2000_area)



plot_dat_2 <- level3 |>  
  distinct(country, .keep_all = T) |>  
  #mutate("nat_area" = rowSums(across(where(is.numeric)), na.rm = T)) %>% 
  select(c(1,13:23))  |>  
  pivot_longer(-1, values_to = "area", names_to = "category") |> 
  filter(category != "Agriculture mosaics with significant natural vegetation") %>% 
  mutate(type_short = str_extract(category, "[A-Z].+")) %>% 
  select(1,3,4) %>% 
  mutate(type_short_level_3 = ifelse(type_short %in% c("Permanently irrigated land", "Annual crops associated with permanent crops", "Non-irrigated arable land", "Rice fields", "Complex cultivation patterns"), "Landbrug i omdrift", type_short)) %>% 
  mutate(type_short_level_3 = ifelse(type_short %in% c("Vineyards", "Fruit trees and berry plantations", "Olive groves", "Agro-forestry areas"), "Agro-forestry og fødevaretræer", type_short_level_3)) %>% 
  mutate(type_short_level_3 = ifelse(type_short == "Pastures", "Permanent græs", type_short_level_3)) %>% 
  filter(country != "Grand Total") %>% 
  left_join(., natura2000_area, by = "country") %>% 
  mutate(proportion_level_3 = area / natura2000_area) %>% 
  group_by(country, type_short_level_3) %>% 
  reframe(proportion_level_3 = proportion_level_3,
          proportion_sum = sum(proportion_level_3, na.rm = T),
          type_short = type_short)
```


# Plotting proportion of agriculture in Natura2000 areas including complex cultivation patterns
```{r}
dark2_colors_max <- c("#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E" )

# Relevel the 'type_short' factor to define the stacking order
order_levels <- c(
  "Annual crops associated with permanent crops",
  "Non-irrigated arable land",
  "Permanently irrigated land",
  "Rice fields",
  "Complex cultivation patterns"
)

median_val <- plot_dat_2 %>% 
  filter(type_short_level_3 == "Landbrug i omdrift") %>% 
  pull(proportion_sum) %>% 
  median() %>% 
  round(., digits = 4)

mean_val <- plot_dat_2 %>% 
  filter(type_short_level_3 == "Landbrug i omdrift") %>% 
  pull(proportion_sum) %>% 
  mean() %>% 
  round(., digits = 4)

plot_dat_2 %>% 
  filter(type_short_level_3 == "Landbrug i omdrift") %>%
  mutate(type_short = factor(type_short, levels = order_levels)) %>%  # Relevel the 'type_short' variable
  ggplot(aes(x = fct_reorder(country, proportion_sum, .desc = TRUE), 
             y = proportion_level_3, 
             fill = type_short)) +
  theme_bw() +
  scale_fill_manual(values = dark2_colors_max, name = "Level 3 Corine land use classes") +  # Use Dark2 palette
  geom_bar(position = "stack", stat = "identity", width = 0.8) +
  theme(axis.title.x = element_blank(),  # Remove x-axis title (label)
        axis.text.x = element_text(angle = 45, hjust = 1),  # Keep x-axis text
        plot.title = element_text(hjust = 0.5),
        legend.background = element_blank(),
        legend.position = c(0.7, 0.78)) +
  scale_y_continuous(expand = c(0.02, 0), labels = scales::percent, limits = c(0, 0.25)) +
  geom_hline(yintercept = median_val, linetype = "dashed", color = "red", linewidth = 0.2) +  # Add stippled line
  geom_hline(yintercept = mean_val, linetype = "dashed", color = "blue", linewidth = 0.2) +  # Add stippled line
  annotate("text", x = 15, y = median_val, 
           label = paste0("EU-Median ", "(", median_val * 100, "%)"), 
           color = "red", hjust = -0.5, vjust = -0.5) +
  annotate("text", x = 15, y = mean_val, 
           label = paste0("EU-Mean ", "(", mean_val * 100, "%)"), 
           color = "blue", hjust = -0.6, vjust = 1.5) +
  ylab("Proportion of Natura 2000 areas")


```

# plotting proportion agriculture excluding complex cultivation patterns
```{r}
dark2_colors_min <- c("#1B9E77", "#D95F02", "#7570B3", "#E7298A")


# excluding complex cultivation patterns and calculating proportions for countries
plot_dat_2_min <- level3 %>% 
  distinct(country, .keep_all = T) %>% 
  #mutate("nat_area" = rowSums(across(where(is.numeric)), na.rm = T)) %>% 
  select(c(1,13:23)) %>% 
  pivot_longer(-1, values_to = "area", names_to = "category") %>% 
  filter(category != "Agriculture mosaics with significant natural vegetation") %>% 
  mutate(type_short = str_extract(category, "[A-Z].+")) %>% 
  select(1,3,4) %>% 
  mutate(type_short_level_3 = ifelse(type_short %in% c("Permanently irrigated land", "Annual crops associated with permanent crops", "Non-irrigated arable land", "Rice fields"), "Landbrug i omdrift", type_short)) %>% 
  mutate(type_short_level_3 = ifelse(type_short %in% c("Vineyards", "Fruit trees and berry plantations", "Olive groves", "Agro-forestry areas"), "Agro-forestry og fødevaretræer", type_short_level_3)) %>% 
  mutate(type_short_level_3 = ifelse(type_short == "Pastures", "Permanent græs", type_short_level_3)) %>% 
  filter(country != "Grand Total") %>% 
  left_join(., natura2000_area, by = "country") %>% 
  mutate(proportion_level_3 = area / natura2000_area) %>% 
  group_by(country, type_short_level_3) %>% 
  reframe(proportion_level_3 = proportion_level_3,
          proportion_sum = sum(proportion_level_3, na.rm = T),
          type_short = type_short)

# calculating median proportion agriculture 
median_val_min <- plot_dat_2_min %>% 
    filter(type_short_level_3 == "Landbrug i omdrift") %>% 
    pull(proportion_sum) %>% 
    median() %>% 
    round(., digits = 4)

# calculating mean proportion agricultiure
mean_val_min <- plot_dat_2_min %>% 
    filter(type_short_level_3 == "Landbrug i omdrift") %>% 
    pull(proportion_sum) %>% 
    mean() %>% 
    round(., digits = 4)

plot_dat_2_min %>% 
    filter(type_short_level_3 == "Landbrug i omdrift") %>% 
    ggplot(aes(x = fct_reorder(country, proportion_sum, .desc = TRUE), y = proportion_level_3, fill = type_short)) +
  theme_bw() +
  scale_fill_manual(values = dark2_colors_min, name = "Level 3 Corine land use classes") +  # Use Set3 palette
  geom_bar(position = "stack", stat = "identity", width = 0.8) +
  theme(axis.title.x = element_blank(),  # Remove x-axis title (label)
        axis.text.x = element_text(angle = 45, hjust = 1),  # Keep x-axis text
        plot.title = element_text(hjust = 0.5),
        legend.background=element_blank(),
        legend.position = c(0.7,0.75)) +
  scale_y_continuous(expand = c(0.02,0), labels = scales::percent, limits = c(0,0.25)) +
  geom_hline(yintercept = median_val_min, linetype = "dashed", color = "red", linewidth = 0.2) +  # Add stippled line
      geom_hline(yintercept = mean_val_min, linetype = "dashed", color = "blue", linewidth = 0.2) +  # Add stippled line

  annotate("text", x = 15, y = median_val_min, label = paste0("EU-Median ", "(",median_val_min*100,"%)"), color = "red", hjust = -0.1, vjust = -0.5) +
    annotate("text", x = 15, y = mean_val_min, label = paste0("EU-Mean ", "(",mean_val_min*100,"%)"), color = "blue", hjust = -0.1, vjust = -0.5)+
  ylab("Proportion of Natura 2000 areas")
```



